---
title: "Images and backends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Images and backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(annotatR)
```

## A handle, not a pixel array

Whole-slide microscopy and hyperspectral cubes are routinely gigabytes in size,
so annotatR never loads them wholesale. Reading an image returns an
`annot_image`: a lightweight, backend-agnostic *handle* that carries the
metadata you need to annotate — dimensions, pyramid levels, bands, pixel size —
together with a lazy accessor that fetches pixels only when asked.

```{r handle}
img <- at_example_image("tissue")
img
dim(img)
```

The `dim()` of a handle is `c(width, height, bands)`. Printing it summarises the
backend that opened it, the data type, and the extent, but nothing above has
touched the pixels themselves. That is the point: you can inspect, transform and
plan annotations against a huge image on a modest machine, because the handle is
a few hundred bytes regardless of how large the underlying file is.

A handle also exposes accessors for everything an annotation workflow needs:
`at_n_bands()` and `at_bands()` for the channels, `at_pixel_size()` for physical
scale, and `at_meta()` for backend-specific keys such as the source path and
data type. Each simply reads a field of the handle — no I/O is involved.

## Pyramid levels

Slide and multiplex formats store the same scene at several resolutions — a
pyramid. Level `0` is full resolution; each higher level is a coarser overview.
The multiplex example is a three-level pyramidal TIFF.

```{r pyramid}
mx <- at_example_image("multiplex")
at_n_levels(mx)
at_is_pyramidal(mx)

levels <- 0:(at_n_levels(mx) - 1L)
data.frame(
  level  = levels,
  width  = vapply(levels, function(l) at_dims(mx, l)[1], integer(1)),
  height = vapply(levels, function(l) at_dims(mx, l)[2], integer(1))
)
```

Levels matter for two reasons. First, an overview lets you display and navigate
the whole image without reading full resolution. Second, an ROI drawn on a
coarse level records the `level` it was created at, so its coordinates can be
rescaled exactly to level 0 (see `at_roi_rescale()` and `at_transform()`). Every
geometry in annotatR is anchored to a pyramid level; the coordinate convention
is image pixels with the origin at the top-left and `y` increasing downward.

## Backends

A *backend* is the reader for a particular family of formats. The registry maps
format names to four functions — read metadata, fetch a tile, detect a file, and
report whether its optional dependencies are installed — so new formats can be
added without touching core code.

```{r backends}
at_backend_list()
```

The `available` column is evaluated lazily: a backend for a format whose support
package is absent still appears in the list but reads nothing until you install
the package. This keeps the package usable with a minimal set of dependencies,
and means a missing optional package never breaks loading — only reading that
particular format. When you call `at_read_image()` without naming a backend, the
registry tries the available backends in priority order and picks the first that
recognises the file; more specific readers are tried before general ones, so an
OME-TIFF is not claimed by the plain raster reader. You can always bypass
detection by passing `backend =` explicitly.

## The universal tile accessor

Whatever the backend, pixels come out through one function, `at_tile()`, and
always as a three-dimensional `[y, x, band]` numeric array — even for a single
band. You choose the pyramid level, an optional pixel window, and an optional
subset of bands.

```{r tile}
tile <- at_tile(mx, level = 0, xrange = c(1, 64), yrange = c(1, 64))
dim(tile)
```

Because the request is scoped, only that 64x64x4 block is read. To display a
cheap thumbnail of the whole image, read the coarsest level in full:

```{r overview}
overview <- at_tile(mx, level = 2)
dim(overview)

dim(at_tile(mx, level = 2, bands = 1))
```

Spectral cubes are handled by the very same accessor; their band metadata simply
carries wavelengths, which you can read back with the image accessors.

```{r spectral}
cu <- at_example_image("cube")
at_is_spectral(cu)
head(at_bands(cu), 3)
range(at_wavelengths(cu))
```

## Registering a custom backend

Adding support for a new format means supplying the four functions and calling
`at_backend_register()`. The `read_fn` returns an `annot_image` — a classed list
of metadata plus a `handle` for the accessor — and the `tile_fn` turns a request
into a `[y, x, band]` array. Here is a complete, self-contained backend for a toy
"constant image" format that stores only a size and a fill value, and synthesises
pixels on demand.

```{r register}
const_read <- function(path, ...) {
  spec <- readRDS(path)                 # list(width, height, bands, value)
  structure(
    list(
      source     = path,
      backend    = "const",
      dims       = c(spec$width, spec$height),
      n_levels   = 1L,
      level_dims = list(c(spec$width, spec$height)),
      n_bands    = spec$bands,
      band_names = paste0("Band ", seq_len(spec$bands)),
      pixel_size = c(1, 1),
      pixel_unit = "px",
      dtype      = "double",
      handle     = list(value = spec$value),   # no pixels, just the fill value
      meta       = list()
    ),
    class = "annot_image"
  )
}

const_tile <- function(img, level, xrange, yrange, bands) {
  nx <- xrange[2] - xrange[1] + 1L
  ny <- yrange[2] - yrange[1] + 1L
  nb <- if (is.null(bands)) img$n_bands else length(bands)
  array(img$handle$value, dim = c(ny, nx, nb))
}

at_backend_register(
  name         = "const",
  read_fn      = const_read,
  tile_fn      = const_tile,
  detect_fn    = function(path) grepl("\\.const$", path, ignore.case = TRUE),
  available_fn = function() TRUE,
  description  = "Constant-value demo backend",
  extensions   = "const"
)
```

The new backend now behaves like any other — it appears in the registry, is
auto-detected from the file, and serves tiles through `at_tile()`.

```{r use-backend}
path <- tempfile(fileext = ".const")
saveRDS(list(width = 64L, height = 48L, bands = 2L, value = 0.5), path)

cimg <- at_read_image(path)
cimg
dim(at_tile(cimg, xrange = c(1, 8), yrange = c(1, 8)))
```

## Memory discipline

The lazy handle is not a convenience but a contract. Everything downstream reads
tile-wise: `at_extract()` walks each ROI's bounding box level by level, and
`at_mask()` rasterises within the region's extent rather than the whole slide. By
requesting only the window and bands you need — as `at_tile()` above does — you
can annotate and extract from images far larger than memory. A well-behaved
backend honours the same discipline: its `tile_fn` should fetch just the
requested window, never the entire level, so that the cost of a call scales with
the tile rather than the image. The `const` backend above is the extreme case —
it holds no pixels at all and synthesises each tile from a single value.

Next, see the *The annotation model* vignette to learn how ROIs, layers and
projects build on these image handles.
