---
title: "Masks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Masks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(annotatR)
```

A *mask* is the artefact that turns hand-drawn regions into something a
segmentation or classification pipeline can consume: an integer raster on the
same pixel grid as the image, where each pixel carries the identity of the
region it belongs to. In annotatR the mask is the core deliverable, and
`at_mask()` is the function that produces it from an `annot_roi`, an
`annot_layer`, or a whole `annot_project`.

## Three mask types

`at_mask()` writes one of three rasters, chosen with `type`:

- **`"binary"`** — a logical matrix: `TRUE` wherever *any* region covers a
  pixel, `FALSE` elsewhere. Use it for foreground/background problems ("is this
  pixel annotated at all?").
- **`"labelled"`** — one distinct positive integer per ROI, assigned in draw
  order. Use it when individual objects must stay separable, e.g. instance
  segmentation where two adjacent tumour nests are different things.
- **`"multiclass"`** — one integer per *label class*, so every ROI sharing a
  label collapses to the same value. Use it for semantic segmentation, where you
  care about the class ("tumour", "stroma") and not the particular ROI.

Every mask is an `annot_mask`: a matrix with an attached `legend` tibble
(`value`, `label`, `layer`, `roi_id`, `n_px`, `colour`) plus `level`, `dims`,
and `mask_type` attributes. Read the legend back with `at_mask_legend()`.

## The pixel-coverage contract

Rasterisation is only reproducible if the pixel/geometry relationship is exact,
so annotatR pins it down. Pixel `(i, j)` covers the half-open square
`[j-1, j) x [i-1, i)`, with `(0, 0)` at the top-left corner of the top-left
pixel; its centre is therefore `(j - 0.5, i - 0.5)`. With the default
`touches = FALSE`, a pixel is in the mask when *its centre* lies inside the
polygon, using a half-open `[lower, upper)` tie rule: a centre on the lower edge
counts as inside, one on the upper edge as outside.

The cleanest way to trust this is to compute a small case by eye. Take a
rectangle from `(2, 2)` to `(5, 5)` on a 10x10 grid:

```{r worked}
r <- at_roi_rect(2, 2, 5, 5, label = "a")
m <- at_mask(r, "binary", dims = c(10, 10))
print(matrix(as.integer(m), 10, 10))
```

Rows and columns 3, 4 and 5 are set, and nothing else. The rectangle spans
`x` in `[2, 5)`; the pixel-centre x-coordinates that fall in that interval are
`2.5, 3.5, 4.5`, i.e. columns 3, 4, 5. The centre at `x = 5.5` (column 6) is
outside, and the lower edge at `x = 2` is inside because centre `2.5 > 2`. The
same reasoning on `y` gives rows 3:5. A four-pixel-wide rectangle covers exactly
nine pixels — no fencepost surprises.

## `touches`: centres versus contact

`touches = TRUE` switches to a "does the polygon touch the pixel at all" rule
(GDAL `ALL_TOUCHED=TRUE`), which grows the region by the fringe pixels a polygon
merely clips. It matters when edges cut through pixel interiors rather than
running along the grid. A rectangle with fractional bounds shows the contrast:

```{r touches}
r2  <- at_roi_rect(2.5, 2.5, 4.5, 4.5, label = "a")
off <- matrix(as.integer(at_mask(r2, "binary", dims = c(8, 8), touches = FALSE)), 8, 8)
on  <- matrix(as.integer(at_mask(r2, "binary", dims = c(8, 8), touches = TRUE)),  8, 8)
off
on
```

With `touches = FALSE` only the pixels whose centres sit inside the rectangle
are kept; with `touches = TRUE` the partially clipped border pixels join too.
Prefer the default for area-faithful masks, and `touches = TRUE` only when you
must not drop thin structures.

## Overlap resolution

When two regions cover the same pixel, `overlap` decides who wins. The ROIs are
rasterised in draw order (ascending `z`, then insertion):

```{r overlap}
lyr <- at_layer("regions")
lyr <- at_layer_add(lyr, at_roi_rect(1, 1, 5, 5, label = "a"))
lyr <- at_layer_add(lyr, at_roi_rect(3, 3, 7, 7, label = "b"))
matrix(as.integer(at_mask(lyr, "labelled", dims = c(8, 8), overlap = "last")),  8, 8)
matrix(as.integer(at_mask(lyr, "labelled", dims = c(8, 8), overlap = "first")), 8, 8)
```

`"last"` (the default) lets the later-drawn region overwrite; `"first"` keeps
the earlier one; `"max"` and `"min"` resolve by pixel value; and `"error"`
aborts the moment any two regions collide, which is the safest choice when
overlaps should never happen.

## Masks from a project

On a real project the mask spans the image grid. The bundled example carries one
layer of three labelled regions:

```{r project}
proj <- at_example_project()
mc <- at_mask(proj, type = "multiclass")
mc
at_mask_legend(mc)
table(as.integer(mc))
```

`at_mask_stats()` summarises each value — pixel counts, area, fraction of the
frame, bounding box and centroid — ready for a report table:

```{r stats}
at_mask_stats(mc)
```

## Writing masks and the legend sidecar

`at_write_mask()` writes an integer TIFF (16-bit by default), and — because a
bare integer raster is meaningless without a key — a companion
`<path>.legend.json` describing it:

```{r write}
dir  <- tempfile("masks-"); dir.create(dir)
path <- file.path(dir, "regions.tif")
at_write_mask(mc, path, overwrite = TRUE)
list.files(dir)
cat(readLines(paste0(path, ".legend.json")), sep = "\n")
```

The schema is self-describing: `annotatR_version`, the `created` timestamp, the
`mask_type`, the pyramid `level`, the raster `dims`, the `source_project`, and
the full `legend` array (one object per value, mirroring the legend tibble).
Any consumer can map pixel values back to labels without touching R.

## Downstream consumption

A Python pipeline reads the TIFF and its legend directly — no R runtime, no
bespoke format:

```python
import json
import tifffile

mask = tifffile.imread("regions.tif")            # integer array, image grid
with open("regions.tif.legend.json") as f:
    legend = json.load(f)

value_to_label = {row["value"]: row["label"] for row in legend["legend"]}
tumour = mask == next(v for v, l in value_to_label.items() if l == "tumour")
```

## Reading masks back

Masks are not a dead end. `at_read_mask()` polygonises an existing mask —
annotatR's own output, or a Cellpose/StarDist/QuPath raster — back into an
editable `annot_layer`, recovering labels from the sidecar when it is present:

```{r read}
back <- at_read_mask(path)
vapply(back$rois, function(r) r$label, character(1))
```

Finally, `at_plot_mask()` renders a mask with its legend for a quick visual
check:

```{r plot}
at_plot_mask(mc)
```

Next, see the **Interoperability** vignette for GeoJSON, QuPath, and CSV
exchange with other tools.
