---
title: "The batch annotation app"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The batch annotation app}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(annotatR)
```

Annotating one image is a job for `at_project()` and the ROI constructors. Annotating a *queue* of images — a plate of serial sections, a batch of slides, a folder of cubes — is what `at_annotate()` is for. It wraps a Shiny and OpenSeadragon canvas around a resumable **session**, so you can work through many images without losing your place, your labels, or your unsaved edits.

## Building a session

A session is a queue of image paths plus a shared annotation template. Build one explicitly with `at_session()`, or grab a ready-made one for experimentation with `at_example_session()`, which copies the bundled tissue image into a temporary directory so the queue has real files to iterate.

```{r session}
sess <- at_example_session(3)
sess
```

Every image starts with status `"pending"`, the cursor sits on the first image, and per-image projects are materialised lazily — nothing is read from disk until you visit it.

## Layer templates and label vocabularies

Two things are shared across the whole queue. The **label vocabulary** is the flat list of allowed labels, bound to the number keys `1`–`9`. The **layer template** is a list of `at_layer()` objects — with styles from `at_style()` — stamped onto every image the first time you open it, so each slide starts with the same empty layers ready to receive ROIs.

```{r template}
regions <- at_layer("regions", labels = c("tumour", "necrosis", "stroma"),
                    style = at_style(colour = "#E64B35", fill_alpha = 0.3))
lesions <- at_layer("lesions", labels = c("focus", "margin"),
                    style = at_style(colour = "#4DBBD5", z = 2))

paths <- at_session_status(sess)$path
templated <- at_session(paths,
                        labels = c("tumour", "necrosis", "stroma", "focus", "margin"),
                        layers = list(regions, lesions),
                        out_dir = tempdir())

at_layers(at_current(templated))
```

## Launching the app

With a session in hand, launch the application. This opens a browser and blocks, so it is never run inside a vignette:

```{r launch, eval=FALSE}
at_annotate(templated)
```

`at_annotate()` is forgiving about its first argument: pass a session, a project, an image, a vector of paths, a directory, or `NULL` to open the bundled example.

## Keyboard shortcuts

The canvas is built for speed, so almost everything has a key. Draw with the tool keys, assign a label with a digit, then commit.

| Key | Action |
|-----|--------|
| `n` / `p` | Next / previous image |
| `N` | Jump to next pending image |
| `1`–`9` | Assign label from the vocabulary |
| `q` | Rectangle tool |
| `w` | Polygon tool |
| `e` | Ellipse / circle tool |
| `r` | Freehand tool |
| `t` | Point tool |
| `y` | Pan / move tool |
| `d` | Delete selected ROI |
| `Ctrl`+`Z` / `Ctrl`+`Shift`+`Z` | Undo / redo |
| `Space` | Toggle mask overlay |
| `Enter` | Commit the current ROI |
| `Shift`+`Enter` | Commit and advance to next image |
| `Shift`+`V` | Copy annotations forward from previous image |
| `f` | Flag image for review |
| `s` | Save session |
| `Ctrl`+`E` | Export |
| `?` | Show help |

## Copy-forward for serial sections

Adjacent serial sections often share most of their anatomy. Rather than redraw it, press `Shift`+`V` to copy every ROI from the previous image onto the current one, then nudge the vertices to fit. Combined with `Shift`+`Enter` (commit and advance), a run of near-identical sections can be annotated in a few keystrokes each.

## Autosave and recovery

By default a session autosaves to `_session.rds` in its `out_dir` on every commit, so a crashed browser or a closed laptop costs you nothing. Save on demand with `at_save_session()` (bound to `s`), and pick up exactly where you left off with `at_resume()`.

```{r resume}
sess <- at_set_status(sess, 1, "complete")
sess <- at_next(sess)

at_save_session(sess)
recovered <- at_resume(file.path(sess$out_dir, "_session.rds"))
recovered
```

The cursor, statuses, materialised projects, and templates all come back intact.

## The session API without the app

Everything the canvas does to a session is available as plain functions, which is what makes batches scriptable and testable. Inspect the queue with `at_session_status()`, walk it with `at_next()` / `at_prev()` / `at_goto()`, and read a richer progress report — one column per label — with `at_manifest()`.

```{r api}
at_session_status(sess)

at_manifest(sess)
```

Use these to seed a session programmatically, pre-fill statuses, or audit progress in a report — no Shiny required.

## Export scoping

Exports (`Ctrl`+`E`) are scoped deliberately. You choose the **layers** to include, the **format** (GeoJSON, QuPath, or a rasterised TIFF mask), and the **breadth** — the current image only, every completed image, or the whole queue. Flagged and skipped images are excluded from batch exports unless you opt them back in, so a "export all complete" pass yields exactly the reviewed annotations and nothing half-finished.

Those export formats — and how annotatR round-trips with QuPath, GeoJSON, and mask TIFFs — are the subject of the next vignette, "Interoperability".
